A general and efficient representation of Ancestral Recombination Graphs

### 2023-10-06 paper.tex
Make Figs 1 & 2 shorter but still readable

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{math,calc,positioning}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

\begin{figure}
\begin{center}
    % FIXME making this 0.75 textwidth for now because it's too tall.
    % https://github.com/tskit-dev/what-is-an-arg-paper/issues/321
    \includegraphics[width=0.75\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
[FIXME Update fig to make wider and less tall, and change $I$ to $D$.
Also possibly add some tabular encoding of gARG data.]
An example genome ARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

@@

\begin{figure}
\centering
\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node [anchor=north west] at (\x1,6) {A};
\node [anchor=north west] at (\x2,6) {B};
\node [anchor=north west] at (\xt,6) {C};

%%% (A) ARG

\node (s0) [greynode] at (\x1 + 0, 0) {};
\node (s1) [greynode] at (\x1 + 3, 0) {};
\node (s2) [greynode] at (\x1 + 6, 0) {};
\node (s3) [greynode] at (\x1 + 3, 1) {};
\node (s4) [greynode] at (\x1 + 1, 2) {};
\node (s5) [greynode] at (\x1 + 5, 3) {};
\node (s6) [greynode] at (\x1 + 3, 4) {};

\draw (s1) -- (s3);
\draw (s0) |- (s4);
\draw (s4) -- (\x1 + 2,2) |- (s3);
\draw (s4) |- (s6);
\draw (s3) -- (\x1 + 4,1) |- (s5);
\draw (s2) |- (s5);
\draw (s5) |- (s6);

%%% (B) Trees
\node (l0) [greynode] at (\x2 + 0, 0) {};
\node (l1) [greynode] at (\x2 + 2, 0) {};
\node (l2) [greynode] at (\x2 + 3, 0) {};
\node (l3) [greynode] at (\x2 + 1, 2) {};
\node (l4) [greynode] at (\x2 + 2, 4) {};

\draw (l0) |- (l3);
\draw (l1) |- (l3);
\draw (l2) |- (l4);
\draw (l3) |- (l4);

\node (r0) [greynode] at (\x3 + 0, 0) {};
\node (r1) [greynode] at (\x3 + 1, 0) {};
\node (r2) [greynode] at (\x3 + 3, 0) {};
\node (r3) [greynode] at (\x3 + 2, 3) {};
\node (r4) [greynode] at (\x3 + 1, 4) {};

\draw (r0) |- (r4);
\draw (r1) |- (r3);
\draw (r2) |- (r3);
\draw (r3) |- (r4);

\foreach \u/\lab in {
        s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$,
        r0/$\textsf{a}$, r1/$\textsf{b}$, r2/$\textsf{c}$,
        l0/$\textsf{a}$, l1/$\textsf{b}$, l2/$\textsf{c}$}
    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};

\foreach \u/\lab in {
        s6/$\textsf{g}$,
        l4/$\textsf{g}$,
        r4/$\textsf{g}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
\foreach \u/\lab in {
        s4/$\textsf{e}$,
        l3/$\textsf{e}$}
    \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {
        s5/$\textsf{f}$,
        r3/$\textsf{f}$}
    \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

\draw[dashed] (\xx,0) -- (\xx, 4);
\node[nodelabel,anchor=north] at (\xx,0) {$x$};

%%% (C) Encoding
\node [nodelabel,anchor=north west] at ($(\xt,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

@@

Note that the \citet{wiuf1999recombination} eARG
in Fig.~\ref{fig-ancestry-resolution} is not particularly
representative, because inference or simulation methods will usually
only generate ARGs containing nodes and edges ancestral to the sample
(see the discussion of the ``Big ARG'' stochastic process in Appendix XXX, however).
Nonetheless, it is an instructive example from the literature which highlights several
important properties of ARGs, and the general point about
the need to resolve ancestral material ``on the fly'' for eARG traversals
holds.

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example genome ARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f} (with breakpoints in red), and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

@@

\begin{figure}
% FIXME add local trees in B
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/320
\centering
% \begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] at (0, 0) {};
\node (s1) [greynode] at (3, 0) {};
\node (s2) [greynode] at (6, 0) {};
\node (s3) [greynode] at (3, 1) {};
\node (s4) [greynode] at (1, 2) {};
\node (s5) [greynode] at (5, 3) {};
\node (s6) [greynode] at (3, 4) {};

\node [anchor=north west] at (0,6) {A};
\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) -- (s3);
\draw (s0) |- (s4);
\draw (s4) -- (2,2) |- (s3);
\draw (s4) |- (s6);
\draw (s3) -- (4,1) |- (s5);
\draw (s2) |- (s5);
\draw (s5) |- (s6);

\node [anchor=north west] at (8,6) {C};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

@@

Note that the \citet{wiuf1999recombination} eARG
in Fig.~\ref{fig-ancestry-resolution} is something of a straw man,
in that typically an inference or backward-time simulation program would only produce an
ARG containing
nodes or edges ancestral to the sample
(see the discussion of the ``Big ARG'' stochastic process in Appendix XXX,
however). Nonetheless, it
is an instructive example from the literature which highlights several
important properties of ARGs, and the general point about
the need to resolve ancestral material ``on the fly'' for eARG traversals
holds.
